FATE summary results

### 2019-09-09 paper draft.Rmd
Office updates

```{r, echo=F}
GOA.biol <- readRDS("GOA_biol_9.4.19.rds") 
GOA.biol$summary
```

@@

new.names <- c("Chignik.coho", "Chignik.pink", "Chiniak.eulachon", "Chiniak.jellyfish", "Chiniak.shrimp", "Cook.Inlet.coho", "Kodiak.coho", "Kodiak.pink", "Pavlof.capelin", "Pavlof.eulachon", "Pavlof.jellyfish", "Pavlof.shrimp", "Prince.William.coho", "Southeast.coho", "Southeast.herring", "Southeast.pink", "South.Peninsula.coho", "South.Peninsula.pink")

@@

ewidata <- as.tibble(read.csv("ewidata.csv", row.names=1))
keep <- grep("AKCLIM_GOA", ewidata$code)
clim.var <- ewidata[keep,]  
keep <- grep("win.sst", clim.var$code)
clim.var <- clim.var[keep,]

@@

@@

set.seed(99)
mod = trend_lm(rotated_modelfit = r, all_years = 1972:2016,
  selected_years = 1972:2016, climate_dat=cli_dat)

names(mod)[2:4] <- c("1972-1988", "1989-2013", "2014-2016")
summary(mod)

@@

melt.mod <- melt(mod, measure.vars=c("1972-1988", "1989-2013", "2014-2016"))
names(melt.mod)[3] <- "era"

pdf("Fig. 3A goa.biol.on.sst.pdf", 6, 4)
ggplot(melt.mod, aes(x=value, color=era)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + labs(title="A) SST")
dev.off()

@@

@@

set.seed(99)
mod2 = trend_lm(rotated_modelfit = r, all_years = 1972:2016,
  selected_years = 1972:2016, climate_dat=cli_dat)

names(mod2)[2:4] <- c("1972-1988", "1989-2013", "2014-2016")

@@

melt.mod <- melt(mod2, measure.vars=c("1972-1988", "1989-2013", "2014-2016"))
names(melt.mod)[3] <- "era"

pdf("Fig. 3B goa.biol.on.pdo.pdf", 6, 4)
ggplot(melt.mod, aes(x=value, color=era)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + labs(title="B) PDO")
dev.off()

@@

And the NPGO.

@@

# 
# npgo <- read.csv("winter.npgo.csv")
# 
# years <- as.numeric(colnames(GOA.biol$best_model$data)) # these are the years for the biology model - 1972:2016
# 
# # set up climate data
# cli_dat <- data.frame(Year=npgo$YEAR,NPGO=npgo$NDJFM)
# 
# cli_dat <- filter(cli_dat, Year %in% years) %>% mutate(NPGO = arm::rescale(NPGO))
# 
# # cli_dat$regime = as.factor(sample(c(0,1,2), size=20, replace=T))
# 
# cli_dat$slope <- 1
# cli_dat$slope[cli_dat$Year %in% 1989:2013] <- 2
# cli_dat$slope[cli_dat$Year %in% 2014:2016] <- 3
# cli_dat$slope <- as.factor(cli_dat$slope)
# 
# # rename rotated
# r <- rotated
# 
# Y = rnorm(length(cli_dat$slope))
# cli_dat = as.data.frame(model.matrix(Y ~ Year + NPGO:slope, data=cli_dat))
# # replace ':' in name for formula
# names(cli_dat) = gsub(":","",names(cli_dat))
# # remove named intercept
# cli_dat = cli_dat[,-which(names(cli_dat)=="(Intercept)")]
# 
# mod2 = trend_lm(rotated_modelfit = r, all_years = 1972:2016,
#   selected_years = 1972:2016, climate_dat=cli_dat)
# 
# 
# summary(mod2)
# 
# names(mod2)[2:4] <- c("1972-1988", "1989-2013", "2014-2016")
# 
# melt.mod <- melt(mod2, measure.vars=c("1972-1988", "1989-2013", "2014-2016"))
# names(melt.mod)[3] <- "era"
# 
# pdf("Fig. 3C goa.biol.on.npgo.pdf", 6, 4)
# ggplot(melt.mod, aes(x=value, color=era)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + labs(title="B) NPGO")
# dev.off()

@@

@@

rotated = rotate_trends(GOA.ich$best_model)

new.names <- c("sandlance", "ronquils", "walleye.pollock", "Pacific.cod", "flathead.sole", "n.rock.sole", "rockfish")

# loadings
pdf("goa.ich.loadings.pdf", 6, 6)
print(plot_loadings(rotated, names = new.names, threshold=0.9))
dev.off()
```

pdf("goa.ich.trend.pdf", 6,6)
print(plot_trends(rotated, years = as.numeric(colnames(Y))))
dev.off()
```

@@

ewidata <- as.tibble(read.csv("ewidata.csv", row.names=1))
keep <- grep("AKCLIM_GOA", ewidata$code)
clim.var <- ewidata[keep,]  
keep <- grep("win.sst", clim.var$code)
clim.var <- clim.var[keep,]

@@

pdf("Fig. x goa.ich.on.sst.pdf", 4, 5)
ggplot(mod, aes(x=SST)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + xlab("Slope")
dev.off()

@@

# loadings
pdf("goa.zoop.loadings.pdf", 6, 6)
print(plot_loadings(rotated, names = new.names, threshold=0.9))
dev.off()
```

pdf("goa.zoop.trend.pdf", 6,6)
print(plot_trends(rotated, years = as.numeric(colnames(Y))))
dev.off()
```

@@

ewidata <- as.tibble(read.csv("ewidata.csv", row.names=1))
keep <- grep("AKCLIM_GOA", ewidata$code)
clim.var <- ewidata[keep,]  
keep <- grep("win.sst", clim.var$code)
clim.var <- clim.var[keep,]

@@

pdf("Fig. x goa.zoop.on.sst.pdf", 4, 5)
ggplot(mod, aes(x=SST)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + xlab("Slope")
dev.off()

>>>>

```{r, echo=F}
GOA.biol <- readRDS("GOA_biol_9.9.19.rds") 
GOA.biol$summary
```

@@

new.names <- c("Chignik.coho", "Chignik.pink", "Chiniak.eulachon", "Chiniak.jellyfish", "Chiniak.P.cod", "Chiniak.shrimp", "Cook.Inlet.coho", "Kodiak.coho", "Kodiak.pink", "Pavlof.capelin", "Pavlof.eulachon", "Pavlof.jellyfish", "Pavlof.P.cod", "Pavlof.shrimp", "Prince.William.coho", "Southeast.coho", "Southeast.herring", "Southeast.pink", "South.Peninsula.coho", "South.Peninsula.pink")

 check <- data.frame(names=names, newnames=new.names)
 check

@@

ewidata <- read.csv("updated goa climate data.csv", row.names=1)

keep <- grep("win.sst", colnames(ewidata))

clim.var <- ewidata[,keep] %>%
  mutate(year=1950:2019) %>%
  gather(key, value, -year)

@@

cli_dat$slope <- 1
cli_dat$slope[cli_dat$Year %in% 1989:2013] <- 2
cli_dat$slope[cli_dat$Year %in% 2014:2017] <- 3
cli_dat$slope <- as.factor(cli_dat$slope)

@@

set.seed(99)
mod = trend_lm(rotated_modelfit = r, all_years = 1972:2017,
  selected_years = 1972:2017, climate_dat=cli_dat)

names(mod)[2:4] <- c("1972-1988", "1989-2013", "2014-2017")
summary(mod)

@@

melt.mod <- melt(mod, measure.vars=c("1972-1988", "1989-2013", "2014-2017"))
names(melt.mod)[3] <- "era"
melt.mod$covar <- "SST"

@@

years <- as.numeric(colnames(GOA.biol$best_model$data)) # these are the years for the biology model - 1972:2017

# set up climate data
cli_dat <- data.frame(Year=pdo$YEAR, PDO=win.pdo)

@@

set.seed(99)
mod2 = trend_lm(rotated_modelfit = r, all_years = 1972:2017,
  selected_years = 1972:2016, climate_dat=cli_dat)

names(mod2)[2:4] <- c("1972-1988", "1989-2013", "2014-2017")

@@

melt.mod2 <- melt(mod2, measure.vars=c("1972-1988", "1989-2013", "2014-2017"))
names(melt.mod2)[3] <- "era"
melt.mod2$covar <- "PDO"

@@

And the NPGO. Leaving this commented out as there is no meaningfull effect of the NPGO in any era!

@@

# npgo <- read.csv("winter.npgo.csv")
# 
# years <- as.numeric(colnames(GOA.biol$best_model$data)) # these are the years for the biology model - 1972:2017
# 
# # set up climate data
# cli_dat <- data.frame(Year=npgo$YEAR,NPGO=npgo$NDJFM)
# 
# cli_dat <- filter(cli_dat, Year %in% years) %>% mutate(NPGO = arm::rescale(NPGO))
# 
# # cli_dat$regime = as.factor(sample(c(0,1,2), size=20, replace=T))
# 
# cli_dat$slope <- 1
# cli_dat$slope[cli_dat$Year %in% 1989:2013] <- 2
# cli_dat$slope[cli_dat$Year %in% 2014:2017] <- 3
# cli_dat$slope <- as.factor(cli_dat$slope)
# 
# # rename rotated
# r <- rotated
# 
# Y = rnorm(length(cli_dat$slope))
# cli_dat = as.data.frame(model.matrix(Y ~ Year + NPGO:slope, data=cli_dat))
# # replace ':' in name for formula
# names(cli_dat) = gsub(":","",names(cli_dat))
# # remove named intercept
# cli_dat = cli_dat[,-which(names(cli_dat)=="(Intercept)")]
# 
# mod3 = trend_lm(rotated_modelfit = r, all_years = 1972:2017,
#   selected_years = 1972:2017, climate_dat=cli_dat)
# 
# 
# summary(mod3)
# 
# names(mod3)[2:4] <- c("1972-1988", "1989-2013", "2014-2017")
# 
# melt.mod3 <- melt(mod3, measure.vars=c("1972-1988", "1989-2013", "2014-2017"))
# names(melt.mod3)[3] <- "era"
# melt.mod3$covar <- "NPGO"
# 
# pdf("Fig. 3C goa.biol.on.npgo.pdf", 6, 4)
# ggplot(melt.mod3, aes(x=value, color=era)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + labs(title="C) NPGO")
# dev.off()

@@

Combine and plot
```{r}
plot.melt <- rbind(melt.mod, melt.mod2)
plot.melt$covar <- reorder(plot.melt$covar, desc(plot.melt$covar))
plot.melt$era <- reorder(plot.melt$era, desc(plot.melt$era))
ggplot(plot.melt, aes(x=era, y=value)) +
  theme_bw() + 
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept=0, lty=2) +
  geom_vline(xintercept = 0, lty=2) +
  facet_wrap(~covar) +
  theme(axis.title.y = element_blank()) +
  ylab("Slope") +
  coord_flip()

ggsave("goa biology era sst pdo slopes.png", width=6, height=4, units="in")
```

@@

rotated = rotate_trends(GOA.ich$best_model) 
new.names <- c("sandlance", "ronquils", "walleye.pollock", "Pacific.cod", "flathead.sole", "n.rock.sole", "rockfish")

 check <- data.frame(names=names, newnames=new.names)
 check

plot.a <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("a) Ichthyoplankton loadings")

plot.a

```{r, echo=F}
# trends
plot.trend <- data.frame(year=as.numeric(as.character(colnames(Y))), trend=as.vector(rotated$trends_mean),
                         lo=as.vector(rotated$trends_lower),
                         hi=as.vector(rotated$trends_upper))

plot.b <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") +
  theme(axis.title.x = element_blank()) + 
  ggtitle("b) Ichthyoplankton trend")

plot.b
# png("combined GOA climate loading and trend plot.png", 8, 4, units = "in", res=300)
# 
# ggpubr::ggarrange(plot.a, plot.b, ncol=2)
# 
# dev.off()
```

@@

```{r, include=F}
#  load sst as the explanatory variable.

ewidata <- read.csv("updated goa climate data.csv", row.names=1)

keep <- grep("win.sst", colnames(ewidata))

clim.var <- ewidata[,keep] %>%
  mutate(year=1950:2019) %>%
  gather(key, value, -year)

@@

ggplot(mod, aes(x=SST)) + geom_density() + geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + xlab("Slope")

@@

plot.d <- ggplot(plot.load, aes(key, value)) +
  theme_bw() +
  geom_violin(fill=cb[3]) +
  geom_hline(yintercept = 0) +
  theme(axis.title.y = element_blank()) + ylab("Loading") +
  coord_flip() +
  ggtitle("d) Zooplankton loadings")

plot.d
```

```{r, echo=F}
# trends
plot.trend <- data.frame(year=as.numeric(as.character(colnames(Y))), trend=as.vector(rotated$trends_mean),
                         lo=as.vector(rotated$trends_lower),
                         hi=as.vector(rotated$trends_upper))

plot.e <- ggplot(plot.trend, aes(year, trend)) +
  theme_bw() +
  geom_line(color=cb[3]) +
  geom_ribbon(aes(ymin=lo, ymax=hi), fill=cb[3], alpha=0.4) +
  geom_hline(yintercept = 0) + ylab("Trend value") +
  theme(axis.title.x = element_blank()) + 
  ggtitle("e) Zooplankton trend")

plot.e
```

@@

```{r, include=F}
#  load sst as the explanatory variable.
ewidata <- read.csv("updated goa climate data.csv", row.names=1)

keep <- grep("win.sst", colnames(ewidata))

clim.var <- ewidata[,keep] %>%
  mutate(year=1950:2019) %>%
  gather(key, value, -year)

@@

set.seed(99)
mod2 = trend_lm(rotated_modelfit = r, all_years = years,
  selected_years = years, climate_dat=cli_dat)

# get 95% CIs
x <- sort(mod2[,2])
print(names(mod2)[2])
print(paste("median: ", median(x)))
print(paste("LCI: ", x[(0.025*25000)]))
print(paste("UCI: ", x[(0.975*25000)]))

mod2$covar <- "SST"

ggplot(mod2, aes(x=covar, y=SST)) + 
  theme_bw() +
  geom_violin() + 
  geom_hline(yintercept=0) + geom_vline(xintercept = 0, lty=2) + xlab("Slope")
